---
title: "Exploratory Analysis for Armed Conflict Data"
author: "Sheharyar Raza"
format: html
editor: visual
---

## Exploratory Data Analysis

This document contains exploratory data analysis for the armed conflict data.

### Table of Contents

1.  First

2.  Second

3.  Third

4.  Fourth

5.  


#### Loading Data
```{r}
#| output: false 
#| echo: false

here() #set wd
# Running the script using source()
df <- read.csv("../data/final_analysis_dataset.csv")


```


#### Initial checks for data dimensions
```{r}
#| output: false 
#| echo: false

df

```


#### Checking all variables for type and structure
```{r}
#| output: false 
#| echo: false

glimpse(df)

```


#### Initial checks for data dimensions
```{r}
#| output: false 
#| echo: false

df

```




#### Initial checks for data dimensions
```{r}
#| output: false 
#| echo: false

summary(df)



```




#### Initial checks for data dimensions
```{r}
#| output: false 
#| echo: false

df

```



```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
